#Doing the Recency Frequency and monetory analysis of ecommerce dataset
```{r}
df=read.csv("d:/Ecommerce Project 2.csv")
head(df)

#displaying the stucture , the datatypes of ecom data frame
str(df)
```
#need to extract numbers from InvoiceNo,Stockcode,CustomerID
#eliminating characters from invoice_no
```{r}
          
inv=gsub('[A-Z]*','',df$InvoiceNo)
inv=as.numeric(inv)
df$InvoiceNo=inv
```

#eliminating characters from CustomerID
```{r}

cid=gsub("[^[:alnum:]]", "",df$CustomerID)
cid=as.numeric(cid)
df$CustomerID=cid

```
```{r}
#reviewing the structure of the data frame again
str(df)


#Analysing the missing values

df_missing=sapply(df,function(x)(sum(is.na(x))))
df_missing[df_missing>0]
# dropping the column x as it has no values
df_missing
df=df[,-9]


#removing the rows with null customer id
df=na.omit(df)
df_missing=sapply(df,function(x)(sum(is.na(x))))
df_missing[df_missing>0]

dim(df)

```

#creating a column total_price as quantity*unit_price

```{r}

df$tot_price=df$Quantity*df$UnitPrice

str(df)

View(df)
```
```{r}

library(plyr)
df_country=count(df$Country)


df_country=df_country[order(df_country$freq),]
head(df_country)

#converting country to factor variables

df$Country=as.factor(df$Country)



```

#creating columns for calculating recency freqiuency and monetory analysis

```{r}

#grouping the  data by customer id ,how much the customer has spend


library(dplyr)

df_price=df%>%
  group_by(CustomerID)%>%
  summarise("tot"=sum(tot_price))
head(df_price)
nrow(df_price)
```

#grouping the  data by customer id ,based on how many times they have transacted
```{r}
library(dplyr)
# df_inv_cid=df %>%
# group_by(CustomerID)%>%
# count(InvoiceNo)



df_txn=df %>%
  group_by(CustomerID) %>%
  summarise(count=n())
head(df_txn)
nrow(df_txn)


```


#separating date into day month and year
```{r}

library(tidyverse)

df2=separate(df,"InvoiceDate",into=c("Day","Month","Year"),sep="-")
head(df2)
unique(df2$Month)
```

#the total transactions by month

```{r}
library(dplyr)
install.packages("dplyr")
df2%>%
  group_by(Month)%>%
  summarise(sum(tot_price))

df2%>%
  group_by(Year)%>%
  summarise(sum(tot_price))

```
#calculating recency of each customer
```{r}

head(df)
df$InvoiceDate=as.Date(df$InvoiceDate,format="%d-%b-%y")
max(df$InvoiceDate)

nrow(df)

library(dplyr)
df_date=df%>%
  group_by(CustomerID)%>%
  summarise(max(InvoiceDate))

nrow(df_date)

df_date$diff_in_days = difftime( max(df$InvoiceDate),df_date$`max(InvoiceDate)`, units = "days")

df_date

```
#combining recency, frequency and monetory parameters in a single data frame
```{r}

rfm_matrix=data.frame(cbind("custid"=df_price$CustomerID,"recency"=df_date$diff_in_days,"frequency"=df_txn$count,"monetory"=df_price$tot))
nrow(rfm_matrix)

head(rfm_matrix)
#install.packages("OneR")
library("OneR")

head(rfm_matrix$monetory)

min(rfm_matrix$monetory)
max(rfm_matrix$monetory)

#creating bins for each variables recency, frequency and monetory

#creating bins for monetory 

summary(rfm_matrix$monetory)
b <- c(-Inf,293.4,648.1,1898.5,1611.7,Inf)
names <- c("1", "2", "3","4","5")
rfm_matrix<-rfm_matrix%>%mutate(mon_bins = cut(monetory, breaks = b,labels=names))
head(rfm_matrix)


#creating bins for recency

summary(rfm_matrix$recency)
b <- c(-Inf,16,50,91.58,143,Inf)
names <- c("5", "4", "3","2","1")
rfm_matrix<-rfm_matrix%>%mutate(rec_bins = cut(recency, breaks = b,labels=names))
head(rfm_matrix)


#creating bins for frequency

summary(rfm_matrix$frequency)
b <- c(-Inf,17,42,93.5,102,Inf)
names <- c("1", "2", "3","4","5")
rfm_matrix<-rfm_matrix%>%mutate(freq_bins = cut(frequency, breaks = b,labels=names))
head(rfm_matrix)


rfm_bin=rfm_matrix[,c(1,5,7,6)]

head(rfm_bin)
str(rfm_bin)
rfm_bin$mon_bins=as.numeric(rfm_bin$mon_bins)
rfm_bin$rec_bins=as.numeric(rfm_bin$rec_bins)
rfm_bin$freq_bins=as.numeric(rfm_bin$freq_bins)
head(rfm_bin)
head(rfm_bin)
str(rfm_bin)

head(rfm_bin)
str(rfm_bin)
rfm_bin$tot_score=(rfm_bin$rec_bins+rfm_bin$freq_bins+rfm_bin$mon_bins)/3
nrow(rfm_bin)
nrow(df)

```


#clustering based on RFM analysis

```{r}
```

#Creating an elbow plot
```{r}

n=1:15
wss=function(k){
  kmod=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],k)
  kmod$tot.withinss
  
}


library(tidyverse)
wss_values=map_dbl(n,wss)
wss_values

plot(n, wss_values,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     col="red",
     ylab="Total within-clusters sum of squares")
```


#making clusters by k=3
```{r}

k_mod3=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=3)
k_mod3$cluster
k_mod3

df_mod3=cbind(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],"cluster"=k_mod3$cluster)
df_mod3$cluster=as.factor(df_mod3$cluster)
k_mod3

head(df_mod3)

head(k_mod3)

```

Accuracy of the cluster with number of clusters 3 =`r k_mod3$betweenss/k_mod3$totss`


#making clusters by k=4
```{r}
head(rfm_bin)

k_mod4=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=4)

df_mod4=cbind(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],"cluster"=k_mod4$cluster)
df_mod4$cluster=as.factor(df_mod4$cluster)

head(df_mod4,20)


head(k_mod4)

```
Accuracy of the cluster with number of clusters 4 `r k_mod4$betweenss/k_mod4$totss`

#making clusters by k=6
```{r}
set.seed(100)

k_mod6=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=6)

df_mod6=cbind(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],"cluster"=k_mod6$cluster)
df_mod6$cluster=as.factor(df_mod6$cluster)

df_mod6$cluster

head(df_mod6)

head(k_mod6)

head(rfm_bin)

```

Accuracy of the cluster with number of clusters 6=`r k_mod6$betweenss/k_mod6$totss`



#scaling 
```{r}

summary(df_mod6)

df_mod6_scale=scale(df_mod6[,c("rec_bins","freq_bins","mon_bins")],center=TRUE,scale=TRUE)

summary(df_mod6_scale)

kmod_61=kmeans(df_mod6_scale,centers=6)
head(kmod_61)

```
Accuracy of the cluster with number of clusters 6 after normalisation=`r kmod_61$betweenss/kmod_61$totss`

# Using the factoextra R package
# The function fviz_cluster() [factoextra package] can be used to easily visualize k-means clusters. It takes k-means results and the original data as arguments. 
#In the resulting plot, observations are represented by points, 
# using principal components if the number of variables is greater than 2. 
#It's also possible to draw concentration ellipse around each cluster.

```{r}
#install.packages("ggpubr")
#install.packages("factoextra")
library(ggpubr)
library(factoextra)
```

#Compute PCA and extract individual coordinates

# Dimension reduction using PCA

```{r}
df_mod6$cluster
res.pca <- prcomp(df_mod6[,c("rec_bins","freq_bins","mon_bins")],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(df_mod6$cluster)
# Data inspection


k_mod6=kmeans(rfm_bin[,c("mon_bins","rec_bins","freq_bins")],centers=6)
df_mod6=cbind(rfm_bin[,c("mon_bins","rec_bins","freq_bins")],"cluster"=k_mod6$cluster)
df_mod6$cluster=as.factor(df_mod6$cluster)
head(df_mod6)
head(k_mod6)

df_mod6$cluster
res.pca <- prcomp(df_mod6[,c("mon_bins","rec_bins","freq_bins")],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(df_mod6$cluster)
```
# Data inspection
```{r}

eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +  stat_mean(aes(color = cluster), size = 4)

```



# Percentage of variance explained by dimensions
```{r}
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)
```

# Visualize k-means clusters
# Color individuals according to the cluster groups
# Add concentration ellipses
# Add cluster centroid using the stat_mean() [ggpubr] R function
```{r}

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
   size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)


library(ggplot2)
head(k_mod6)
head(df_mod6)

library(plotly)

head(df_mod6)

nrow(df_mod6)

df_mod6$cluster

plot_ly(x=df_mod6$rec_bins, y=df_mod6$freq_bins, z=df_mod6$mon_bins, type="scatter3d", mode="markers", color=df_mod6$cluster)


head(df_mod6,15)

df_cluster1=df_mod6[df_mod6==1,]
nrow(df_cluster1)
df_cluster2=df_mod6[df_mod6==2,]
nrow(df_cluster2)
df_cluster3=df_mod6[df_mod6==3,]
nrow(df_cluster3)
df_cluster4=df_mod6[df_mod6==4,]
nrow(df_cluster4)
df_cluster5=df_mod6[df_mod6==5,]
nrow(df_cluster5)
df_cluster6=df_mod6[df_mod6==6,]
nrow(df_cluster6)
```





#hierarchical clustering


#scaling the variables in hierarchical clustering
```{r}

df_scale=scale(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],center=TRUE,scale=TRUE)


summary(df_scale)
dist_mat <- dist(df_scale, method = 'euclidean')
```
#plotting dendrogram
```{r}

hclust_avg <- hclust(dist_mat, method = 'ward.D2')
plot(hclust_avg)
```

#creating the required number of clusters
```{r}
member = cutree(hclust_avg,6)

h_clust=rfm_bin
h_clust$cluster=member
```


#plotting the clusters formed using hierarchical clustering
```{r}
library(cluster)
res.pca <- prcomp(h_clust[,c("mon_bins","rec_bins","freq_bins")],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(h_clust$cluster)
# Data inspection



ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

#shiloutte analysis

```{r}

df6_scale=data.frame(scale(rfm_bin[,c("mon_bins","rec_bins","freq_bins")],center=TRUE,scale=TRUE))
k_mod6_1=kmeans(df6_scale,centers=6,nstart=25)
k_mod6_1

head(df6_scale)
df6_scale$cluster=k_mod6_1$cluster

head(df6_scale)

library(cluster)

s =silhouette(df6_scale$cluster, dist(df6_scale[,-4],method = 'euclidean'))

plot(s, col=1:3, border=NA)
```

# the average silhoutte width is .43 ,more it is towards 1 the better

#Silhoutte analysis for hierarchical clustering
```{r}


head(h_clust)

s =silhouette(h_clust$cluster, dist(df6_scale,method = 'euclidean'))
plot(s, col=1:3, border=NA)
```

#the average silhoutte width is .33 
# which is lesser thaan kmeans

#Increasing the number of clusters further
```{r}

k_mod8=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=8)

df_mod8=cbind(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],"cluster"=k_mod8$cluster)
df_mod8$cluster=as.factor(df_mod8$cluster)

df_mod8$cluster

head(k_mod8)
```
Accuracy of the cluster with number of clusters 8`r k_mod8$betweenss/k_mod8$totss`



```{r}

k_mod9=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=9)

df_mod9=cbind(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],"cluster"=k_mod9$cluster)
df_mod9$cluster=as.factor(df_mod9$cluster)

df_mod9$cluster

head(k_mod9)
```

# Within cluster sum of squares by cluster:
#   [1] 340.9895 117.6919 386.8711 687.8474 926.0398 238.7811 989.9324 343.0778 109.8140
# (between_SS / total_SS =  85.6 %)
```{r}


k_mod10=kmeans(rfm_bin[,c("rec_bins","freq_bins","mon_bins")],centers=10)

df_mod10=cbind(rfm_bin[,c("custid","rec_bins","freq_bins","mon_bins")],"cluster"=k_mod10$cluster)
df_mod10$cluster=as.factor(df_mod10$cluster)

head(df_mod10)

head(k_mod10)

```
#customer segmentation by clustering into 10 clusters
#recency high(new customers) ,frequency low ,monetory all types
```{r}
df_cluster1=df_mod10[df_mod10$cluster==1,] #(x4-5x) high recency

head(df_cluster1,20)  
str(df_cluster1)

library(plyr)
# count(df_cluster1$rec_bins)
# count(df_cluster1$freq_bins)
# count(df_cluster1$mon_bins)
nrow(df_cluster1)
```

#cluster 2 frequency high,recency high, monetory moderate to high,best customers

```{r}

df_cluster2=df_mod10[df_mod10$cluster==2,] 
tail(df_cluster2,20) 
nrow(df_cluster2)  
# count(df_cluster2$rec_bins)
# count(df_cluster2$freq_bins)
# count(df_cluster2$mon_bins)
```
#cluster 3 average customers , low to average recency, frequency and spending capability
```{r}
df_cluster3=df_mod10[df_mod10$cluster==3,]
nrow(df_cluster3)   
# count(df_cluster3$rec_bins)
# count(df_cluster3$freq_bins)
# count(df_cluster3$mon_bins)

View(df_cluster3) 
```
#cluster 4 recent buying customers,new
```{r}


df_cluster4=df_mod10[df_mod10$cluster==4,]
nrow(df_cluster4)

head(df_cluster4,15) 
# count(df_cluster4$rec_bins)
# count(df_cluster4$freq_bins)
# count(df_cluster4$mon_bins)
```
#cluster 5 frequently buying customers loyal
```{r}



df_cluster5=df_mod10[df_mod10$cluster==5,]
nrow(df_cluster5)

# count(df_cluster5$rec_bins)
# count(df_cluster5$freq_bins)
# count(df_cluster5$mon_bins)


tail(df_cluster5,15) 
```
#cluster 6 high paying customers having high frequency of buying
```{r}
df_cluster6=df_mod10[df_mod10$cluster==6,]



tail(df_cluster6,15) 
nrow(df_cluster6)


# count(df_cluster6$rec_bins)
# count(df_cluster6$freq_bins)
# count(df_cluster6$mon_bins)

```
#cluster 7 slipping ....high paying customers having high frequency of buying  in the past but dont buy now
```{r}
df_cluster7=df_mod10[df_mod10$cluster==7,]
tail(df_cluster7,15) 
nrow(df_cluster6)


# count(df_cluster7$rec_bins)
# count(df_cluster7$freq_bins)
# count(df_cluster7$mon_bins)
```





#cluster 8 low frequency customers who rarely buy but consist of low to high spenders

```{r}
df_cluster8=df_mod10[df_mod10$cluster==8,]
tail(df_cluster8,15) 
nrow(df_cluster8)


# count(df_cluster8$rec_bins)
# count(df_cluster8$freq_bins)
# count(df_cluster8$mon_bins)
```


#cluster 9 recent  buyers frquent buyers, high spenders ,best customers
```{r}
df_cluster9=df_mod10[df_mod10$cluster==9,]

tail(df_cluster9,15) 
nrow(df_cluster9)


# count(df_cluster9$rec_bins)
# count(df_cluster9$freq_bins)
# count(df_cluster9$mon_bins)
```

# cluster 10 low spenders, frequent buyers
```{r}
df_cluster10=df_mod10[df_mod10$cluster==10,]
tail(df_cluster10,15) 
nrow(df_cluster10)


# count(df_cluster10$rec_bins)
# count(df_cluster10$freq_bins)
# count(df_cluster10$mon_bins)

```


#high valued customers have RFM score of 5,5,5
```{r}
df_mod10$customerID=rfm_bin$custid

df_mod10[ df_mod6$rec_bins==5 & df_mod6$freq_bins==5 & df_mod6$mon_bins==5 ,]


```
#plot with 10 clusters
```{r}

res.pca <- prcomp(df_mod10[,c("mon_bins","rec_bins","freq_bins")],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(df_mod10$cluster)
ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```



































